A Typed Programmatic Interface to Contracts on the Blockchain

### 2021-02-19 papers/execution-model/model.tex
some cutting

\subsection{Contributions}
\label{sec:contributions}

\begin{itemize}
\item An operational semantics for functional programs running
  alongside smart contracts in a blockchain. 
\item Proofs of various properties of the combined system.
\item An implementation of a low-level OCaml-API to the Tezos
  blockchain, which corresponds to the operational semantics.
\end{itemize}

@@

To implement the bidding strategy, we first check the state of the
contract to find the current highest bid. As long as bidding is
allowed and the current bid is below our \lstinline/limit/, we update
our bid by a given amount \lstinline/step/, and then keep watching the state of the contract by
polling it every five minutes. 
\begin{lstlisting}[numbers=none,basicstyle=\small]
let rec poll (limit, step, ((bidding, _), (high_bid, _))) =
  if bidding and high_bid < limit then
    Cl.call_contract (contract, 
      right (min (high_bid + step, limit))); (* entrypoint %bid *)
    Time.sleep(5 * 60);
    let state = Cl.get_state (contract);
    poll (limit, step, state)
in
poll (limit, step, Cl.get_state (contract))
\end{lstlisting}
The \lstinline{get_state} function queries data on
the blockchain and returns the storage of a contract.

@@

In summary, a useful smart-contract-API  has facilities to
\begin{itemize}
\item query the current state of the blockchain (e.g., fees in the
  current block),
\item query the state and the balance of a contract (e.g., to obtain
  the current highest bid),
\item originate contracts, invoke contracts, and initiate
  transfers. To this end, the program using the API has to run on
  behalf of some authority (i.e., expressed by holding the private key
  for some account).
\end{itemize}
All these facilities are supported by the (untyped) RPC interface of the Tezos
blockchain, which is the basis of our implementation.

\section{Execution Model}
\label{sec:execution-model}
The context of our work is the Tezos blockchain \cite{tezos-whitepaper, tezos-intropaper}. Tezos is a self-amending blockchain that improves several aspects compared to established 
blockchains. Tezos introduces an original consensus algorithm called Liquid Proof of Stake that allows consensus not only on the state of its ledger, like Bitcoin \cite{bitcoin-whitepaper} or Ethereum \cite{eth-whitepaper}, but also on how the economic protocol and the nodes should  upgrade, and serves as a decentralized smart contract platform that focuses on formal methods to improve security.

@@

\begin{figure}[tp]
  \begin{align*}
    \LET\ \mathit{oph} &=  \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
         & \FIX \lambda f.\lambda x. \MATCH\ \GETSTATUS{\mathit{oph}}\ \WITH \\
         & \mid \STATUSPENDING \to \text{Time.sleep(1)}; f () \textrm{\qquad(* not yet included,
           continue waiting *)} \\
         & \mid \STATUSINCLUDING (n) \to \text{if } n < 50\text{ then
           Time.sleep(1)};f ()\textrm{\quad(* included with $n$
           blocks confirmation *)}\\
         & \mid \STATUSTIMEOUT \to \textrm{(* lost operation, try
           again *)}
  \end{align*}
  \caption{Pseudocode example for a token transfer}
  \label{fig:pseudocode-example}
\end{figure}
Figure~\ref{fig:pseudocode-example} shows an incomplete example
expression that transfers tokens from $\PUK$ to $\PUK'$ and then
checks the status of the operation. The outer loop to raise the fee
and retry is omitted.

Our type system does not include subtyping. The relation in
Figure~\ref{fig:subtyping} just etablishes pairs of types that can be
subjected to casting. A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType \TYPEU$ (upcast) or $\TYPEU \SubType \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter type.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\TPUH \SubType \TADDR}

    \inferrule{}{\TPUK \SubType \TADDR}

    \inferrule{}{\TCONTRACT\ \TYPE \SubType \TPUH}
  \end{mathpar}
  \caption{Subtyping}
  \label{fig:subtyping}
\end{figure}
Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified, and $\JTypeValue\STRING\TYPE$ for serialized values as
stored on the blockchain. The latter judgment states $\STRING$ is
a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}

@@

    % \inferrule{}{
    % \EC{\FST (\VAL_1, \VAL_2)} \ExprTrans \EC{\VAL_1}
    % }
  %   
    %   \inferrule{}{
    %   \EC{\SND (\VAL_1, \VAL_2)} \ExprTrans \EC{\VAL_2}
    % }
  %   
    \inferrule{ \TRY \notin \EC[F]{} }{ \EC{\TRY\ {\EC[F]{\RAISE\
            \VAL}}\ \EXCEPT\ \EXPR} \ExprTrans \EC{\EXPR\ \VAL} }

    \inferrule{\TYPE \SubType \TYPEU}{ \EC{\CAST\VAL\TYPE\TYPEU} \ExprTrans \EC{\VAL}}
    %
    % \inferrule{}{ \EC{\MATCH\ (\VAL_1, \VAL_2) \WITH\ (x_1,
    %     x_2)\to\EXPR \mid \dots} \ExprTrans \EC{\EXPR[\VAL_1,
    %     \VAL_2/x_1, x_2]} }
    % \inferrule{}{ \EC{\MATCH\ \NIL\ \WITH\ \NIL\to\EXPR \mid \dots}
    %   \ExprTrans \EC{\EXPR} }
    % \inferrule{}{ \EC{\MATCH\ (\CONS\ \VAL_1\ \VAL_2)\ \WITH\ (\CONS\
    %     x_1\ x_2)\to\EXPR \mid \dots} \ExprTrans
    %   \EC{\EXPR[\VAL_1,\VAL_2/x_1, x_2]} }
    % \inferrule{}{ \EC{\MATCH\ \NONE\ \WITH\ \NONE\to\EXPR \mid \dots}
    %   \ExprTrans \EC{\EXPR} }
    % \inferrule{}{ \EC{\MATCH\ (\SOME\ \VAL)\ \WITH\ (\SOME\ x)\to\EXPR
    %     \mid \dots} \ExprTrans \EC{\EXPR[\VAL/x]} }
    % \inferrule{}{ \EC{\MATCH\ (\LEFT\ \VAL)\ \WITH\ (\LEFT\ x)\to\EXPR
    %     \mid \dots} \ExprTrans \EC{\EXPR[\VAL/x]} }
    % \inferrule{}{ \EC{\MATCH\ (\RIGHT\ \VAL)\ \WITH\ (\RIGHT\
    %     x)\to\EXPR \mid \dots} \ExprTrans \EC{\EXPR[\VAL/x]} }
  \end{mathpar}
  \caption{Select expression reduction rules (pattern matching omitted)}
  \label{fig:expression-reduction}
\end{figure}
The internal transitions of a node are just evaluation of expressions.
\begin{mathpar}
  \inferrule[Node-Eval]
  {
    \EXPR \ExprTrans \EXPR'
  }{
    [\EC\EXPR :: \EXPRS, \ACCOUNTS] \NodeTrans{}
    [\EC{\EXPR'} :: \EXPRS, \ACCOUNTS]
  }
\end{mathpar}
The remaining transitions affect a local node in the context of the
blockchain. To this end, any local node may be selected.
\begin{mathpar}
  \inferrule[Config-System]{
    \NODE\|\BLOCKCHAIN \SystemTrans \NODE'\|\BLOCKCHAIN'
}{
    \BLOCKCHAIN[\NODE :: \overline{\NODE}] \SystemTrans
    \BLOCKCHAIN'[\NODE' :: \overline{\NODE}]
  }

@@

>>>>

\subsubsection{Contributions}
\label{sec:contributions}

\begin{itemize}
\item An operational semantics for functional programs running
  alongside smart contracts in a blockchain. 
\item Established various properties of the combined system withs
  proofs in upcoming techreport.
\item An implementation of a low-level OCaml-API to the Tezos
  blockchain, which corresponds to the operational semantics.
\end{itemize}

@@

To implement the bidding strategy, we first check the state of the
contract to find the current highest bid. As long as bidding is
allowed and the current bid is below our \lstinline/limit/, we update
our bid by a given amount \lstinline/step/, and then keep watching the state of the contract by
polling it every five minutes. 
\begin{lstlisting}[numbers=none,basicstyle=\small]
let rec poll (limit, step, ((bidding, _), (high_bid, _))) =
  if bidding and high_bid < limit then
    Cl.call_contract (contract, 
      right (min (high_bid + step, limit))); (* entrypoint %bid *)
    Time.sleep(5 * 60);
    let state = Cl.get_state (contract);
    poll (limit, step, state)
in
poll (limit, step, Cl.get_state (contract))
\end{lstlisting}
The \lstinline{get_state} function obtains the storage of a contract from
the blockchain.

@@

In summary, a useful smart-contract-API  has facilities to
\begin{itemize}
\item query the current state of the blockchain (e.g., fees in the
  current block),
\item query storage and balance of a contract (to obtain
  the current highest bid),
\item originate contracts, invoke contracts, and initiate
  transfers. Hence, the API has to run on
  behalf of some account (by holding its private key).
\end{itemize}
All these facilities are supported by the (untyped) RPC interface of the Tezos
blockchain, which is the basis of our implementation.

\section{Execution Model}
\label{sec:execution-model}
The context of our work is the Tezos blockchain \cite{tezos-whitepaper,tezos-intropaper}. Tezos is a self-amending blockchain that improves several aspects compared to established 
blockchains. Tezos introduces an original consensus algorithm called Liquid Proof of Stake that allows consensus not only on the state of its ledger, like Bitcoin \cite{bitcoin-whitepaper} or Ethereum \cite{eth-whitepaper}, but also on how the economic protocol and the nodes should  upgrade, and serves as a decentralized smart contract platform that focuses on formal methods to improve security.

@@

% \begin{figure}[tp]
%   \begin{align*}
%     \LET\ \mathit{oph} &=  \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
%          & \FIX \lambda f.\lambda x. \MATCH\ \GETSTATUS{\mathit{oph}}\ \WITH \\
%          & \mid \STATUSPENDING \to \text{Time.sleep(1)}; f () \textrm{\qquad(* not yet included,
%            continue waiting *)} \\
%          & \mid \STATUSINCLUDING (n) \to \text{if } n < 50\text{ then
%            Time.sleep(1)};f ()\textrm{\quad(* included with $n$
%            blocks confirmation *)}\\
%          & \mid \STATUSTIMEOUT \to \textrm{(* lost operation, try
%            again *)}
%   \end{align*}
%   \caption{Pseudocode example for a token transfer}
%   \label{fig:pseudocode-example}
% \end{figure}
% Figure~\ref{fig:pseudocode-example} shows an incomplete example
% expression that transfers tokens from $\PUK$ to $\PUK'$ and then
% checks the status of the operation. The outer loop to raise the fee
% and retry is omitted.

We define a relation $\SubType$ that establishes pairs of types that can be
subjected to casting. It is given by the axioms
$\inferrule{}{\TPUH \SubType \TADDR}$, $\inferrule{}{\TPUK \SubType
  \TADDR}$, and $\inferrule{}{\TCONTRACT\ \TYPE \SubType \TPUH}$.
A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType \TYPEU$ (upcast) or $\TYPEU \SubType \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter type.

Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified, and $\JTypeValue\STRING\TYPE$ for serialized values as
stored on the blockchain. The latter judgment states $\STRING$ is
a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}

@@

    \inferrule{}{ \EC{\TRY\ \VAL\ \EXCEPT\ \EXPR} \ExprTrans \EC{\VAL} }

    % \inferrule{}{
    % \EC{\FST (\VAL_1, \VAL_2)} \ExprTrans \EC{\VAL_1}
    % }
  %   
    %   \inferrule{}{
    %   \EC{\SND (\VAL_1, \VAL_2)} \ExprTrans \EC{\VAL_2}
    % }
  %   
    \inferrule{\TYPE \SubType \TYPEU}{ \EC{\CAST\VAL\TYPE\TYPEU} \ExprTrans \EC{\VAL}}

    \inferrule{ \TRY \notin \EC[F]{} }{ \EC{\TRY\ {\EC[F]{\RAISE\
            \VAL}}\ \EXCEPT\ \EXPR} \ExprTrans \EC{\EXPR\ \VAL} }
    %
    % \inferrule{}{ \EC{\MATCH\ (\VAL_1, \VAL_2) \WITH\ (x_1,
    %     x_2)\to\EXPR \mid \dots} \ExprTrans \EC{\EXPR[\VAL_1,
    %     \VAL_2/x_1, x_2]} }
    % \inferrule{}{ \EC{\MATCH\ \NIL\ \WITH\ \NIL\to\EXPR \mid \dots}
    %   \ExprTrans \EC{\EXPR} }
    % \inferrule{}{ \EC{\MATCH\ (\CONS\ \VAL_1\ \VAL_2)\ \WITH\ (\CONS\
    %     x_1\ x_2)\to\EXPR \mid \dots} \ExprTrans
    %   \EC{\EXPR[\VAL_1,\VAL_2/x_1, x_2]} }
    % \inferrule{}{ \EC{\MATCH\ \NONE\ \WITH\ \NONE\to\EXPR \mid \dots}
    %   \ExprTrans \EC{\EXPR} }
    % \inferrule{}{ \EC{\MATCH\ (\SOME\ \VAL)\ \WITH\ (\SOME\ x)\to\EXPR
    %     \mid \dots} \ExprTrans \EC{\EXPR[\VAL/x]} }
    % \inferrule{}{ \EC{\MATCH\ (\LEFT\ \VAL)\ \WITH\ (\LEFT\ x)\to\EXPR
    %     \mid \dots} \ExprTrans \EC{\EXPR[\VAL/x]} }
    % \inferrule{}{ \EC{\MATCH\ (\RIGHT\ \VAL)\ \WITH\ (\RIGHT\
    %     x)\to\EXPR \mid \dots} \ExprTrans \EC{\EXPR[\VAL/x]} }
  \end{mathpar}
  \caption{Select expression reduction rules (pattern matching omitted)}
  \label{fig:expression-reduction}
\end{figure}
The internal transitions of a node are just evaluation of expressions.
\begin{mathpar}
  \inferrule[Node-Eval]
  {
    \EXPR \ExprTrans \EXPR'
  }{
    [\EC\EXPR :: \EXPRS, \ACCOUNTS] \NodeTrans{}
    [\EC{\EXPR'} :: \EXPRS, \ACCOUNTS]
  }
\end{mathpar}
The remaining transitions affect a local node in the context of the
blockchain. To this end, any local node may be selected.
\begin{mathpar}
  \inferrule[Config-System]{
    \NODE\|\BLOCKCHAIN \SystemTrans \NODE'\|\BLOCKCHAIN'
}{
    \BLOCKCHAIN[\NODE :: \overline{\NODE}] \SystemTrans
    \BLOCKCHAIN'[\NODE' :: \overline{\NODE}]
  }

@@

\newpage{}
